### 2015-05-15 whoop.tex
Minor fix.

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity than the level of pairs. However, a pair-wise analysis can be trivially parallelized for performance, although we leave this for future work.}  The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged to an automated theorem prover.

@@

\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may start executing at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.\footnote{We do \emph{not} consider the case where one thread spawns another thread, which does not occur in the context of drivers; rather we aim to capture the scenario where additional threads are launched by the environment.}  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of \emph{shared} variables $\sharedvars$, a finite set of mutexes $\mutexes$, and a finite set of \emph{thread templates}.  A thread template $T$ consists of a finite set of procedures $\procedures{T}$ and a finite set of private variables $\variables{T}$.  A designated procedure $\mainforthread{T} \in \procedures{T}$ denotes the starting point for execution of $T$ by a thread.  Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.  A basic block either has a single successor or a pair of successors.  In the latter case, an \emph{exit condition} over thread-private variables determines the successor to which control should flow on block exit.

The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

\begin{figure}
\footnotesize
\center
\begin{tabular}{lp{5.5cm}}
\textbf{Statement Form} & \textbf{Notes} \\
\toprule

@@

\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.  Each thread consists of an id, drawn from $\ids$, a thread template $T$ an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

\noindent\textbf{Lockset Instrumentation }
%
For templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we want to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.
%
To this end, we first \emph{instrument} a template $T$ for lockset analysis (see \S\ref{bg:lockset}).  Given an arbitrary symbol $i$, we define the instrumentation of $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.

Renaming is straightforward: all occurrences of each private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $T$ and $U$ are the same.

\begin{figure}
\footnotesize
\center
\begin{tabular}{ll}
\textbf{Original Statement} & \textbf{Instrumented Statement} \\
\toprule

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset $\currentlockset{i} \subseteq \powerset{\mutexes}$ to record the mutexes currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset $\lockset{s}{i}$ to record the mutexes that are consistently held when the thread accesses $s$.
%
The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.  For a shared variable assignment $s = e$, we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).  A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating $\currentlockset{i}$ to add and remove mutex $m$, respectively.

\noindent\textbf{Shared State Abstraction }
%
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.  The instrumentation of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in this fashion, it is possible to completely dispense with the shared variables after the lockset instrumentation has been performed.  As a result, when instrumenting a shared variable write, the effect of the write is not explicitly modeled.

\noindent\textbf{Sequentialisation }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

\begin{figure}
\footnotesize
\begin{tabular}{l}
$\currentlockset{i} = \emptyset;$ $\hasbeenread{i} = \emptyset;$ $\hasbeenwritten{i} = \emptyset;$ \\
$\currentlockset{j} = \emptyset;$ $\hasbeenread{j} = \emptyset;$ $\hasbeenwritten{j} = \emptyset;$ \\
\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \medskip
\\

@@

\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct, it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate, the assertion checked at the end of the program guarantees at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.

@@

\noindent\textbf{Architecture }
%
Figure~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).  For the remaining components, we were able to reuse industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based tool that traverses the abstract syntax tree (AST) of the original driver and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions. Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is parsed by \whoop to be used during the instrumentation.

\begin{figure}[t]
\begin{lstlisting}
const struct file_operations nvram_fops = {
  .llseek = nvram_llseek,
  .read = read_nvram,
  .write = write_nvram,
  .unlocked_ioctl = nvram_unlocked_ioctl
};
\end{lstlisting}
\caption{Entry points definitions in the generic\_nvram driver}
\label{fig:entrypoints}
\end{figure}

\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.
%
SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse robust existing techniques and focus instead on verification efforts.

To achieve scalability, SMACK uses a \emph{split} memory model that exploits an alias analysis to soundly partition memory locations into non-overlapping equivalence classes that do not alias. This has been shown to lead to more tractable verification compared with a \emph{monolithic} model where the heap is considered to be an array of bytes~\cite{rakamaric2009scalable}. The split memory model is based on \emph{memory regions}, which are maps of integers that model the heap. A benefit of using this model is that distinct memory regions denote disjoint sections of the heap. We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}.

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite{corbet2005linux}.

\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop, we experimented with analyzing recursion-free drivers using full inlining.  We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined.

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.

Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional entry points by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and allows our analysis to scale to large drivers.  The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

Symbolic verification of a pair of entry points works by (i) instrumenting each entry point with additional state to record locksets, and (ii) attempting to verify a sequential program that executes the instrumented entry points in sequence and then asserts, for each shared location, that the locksets for each entry point with respect to this location have a non-empty intersection.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions~\cite{barnett2005weakest} (VCs) are generated and discharged by an automated theorem prover.

@@

\medskip\noindent\textbf{Concurrent Programming Model }
%
We consider a concurrent programming model where an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  In the context of device drivers, a thread template corresponds to a driver entry point, and multiple instances of the same thread template may execute concurrently, just as multiple invocations of a single driver entry point may be concurrent.  Further threads may be launched at any point during execution; in the context of device drivers this corresponds to the OS invoking additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in \S\ref{sec:implementation} we discuss interesting practical issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of \emph{thread templates}.  A thread template, $T$, consists of a finite set of procedures, $\procedures{T}$, and a finite set of private variables, $\variables{T}$.  A designated procedure, $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of $T$ by a thread.  Each procedure of $\procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.  A basic block either has a single successor or a pair of successors.  In the latter case, an \emph{exit condition} over thread-private variables determines the successor to which control should flow on block exit.

The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation, below, and a program that does not satisfy this restriction can be trivially pre-processed into one that does via the introduction of additional private temporary variables to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

\begin{figure}
\begin{tabular}{lp{5.5cm}}
\textbf{Statement Form} & \textbf{Notes} \\
\toprule

@@

\medskip\noindent\textbf{Semantics }
%
Let $\ids$ be an infinite set from which dynamic thread ids will be drawn.  The state of a running concurrent program consists of: a valuation of the shared variables, $\sharedvars$; a mapping that associates each mutex in $\mutexes$ with an id from $\ids$, recording which thread currently holds the mutex, or with a special value $\bot \notin \ids$ to indicate that the mutex is not held by any thread; and a list of \emph{threads}.  Each thread consists of an id, drawn from $\ids$, a thread template, $T$, an index indicating the next statement of $T$ to be executed by the thread, and a valuation of the thread private variables, $\variables{T}$.  If multiple threads are instances of the same template $T$, then each thread carries a \emph{separate} valuation of the private variables for this template.

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

\medskip\noindent\textbf{Lockset Instrumentation }
%
For thread templates $T$ and $U$ (including the possibility that $T$ and $U$ are equal) we wish to check whether it is possible for a thread executing $T$ to race with a thread executing $U$, in the presence of arbitrarily many further concurrently executing threads.

To achieve this, we first \emph{instrument} a thread template for lockset analysis (see \S\ref{bg:lockset}).  Given an arbitrary symbol, $i$, we define the instrumentation of a template $T$ with respect to $i$, denoted $\instrument{T}{i}$.  There are two aspects to this instrumentation phase: \emph{renaming} and \emph{lockset instrumentation}.

Renaming is straightforward: all occurrences of private variable $x \in \variables{T}$ used in $T$ are replaced with a renamed variable $\instrument{x}{i}$ in $\instrument{T}{i}$, and every procedure $f \in \procedures{T}$ is renamed (both at its declaration site and at all call sites) to $\instrument{f}{i}$ in $\instrument{T}{i}$.  The purpose of renaming is to ensure that when we analyze a pair of templates, $T$ and $U$, both templates execute distinct procedures and operate on distinct private data.  This is vital in the case where $T$ and $U$ are the same.

\begin{figure}
\begin{tabular}{ll}
\textbf{Original Statement} & \textbf{Instrumented Statement} \\
\toprule

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.

The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.  This transformation is described in Figure~\ref{fig:instrumentation}, where for an expression $e$ we use $\instrument{e}{i}$ to denote $e$ after renaming with respect to $i$.  A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).  A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.

\medskip\noindent\textbf{Shared State Abstraction }
%
Recall that while our aim is to perform race analysis for pairs of threads, we must be sure to account for possible side-effects due to other threads that are running concurrently.  The instrumentation of Figure~\ref{fig:instrumentation} achieves this via \emph{nondeterminism}: when reading from a shared variable $s$, a nondeterministic value is returned.  This is reflected by the use of a $\havockeyword$ command, which sets its argument to an arbitrary value.  Because all shared state accesses are abstracted in this fashion, it is possible to completely dispense with the shared variables after the lockset instrumentation has been performed.  As a result, when instrumenting a shared variable write, the effect of the write is not explicitly modeled.

\medskip\noindent\textbf{Sequentialisation }
%
The pseudocode of Figure~\ref{fig:sequentialization} shows the sequential program that we analyze in order to prove race-freedom for a pair of thread templates $T$ and $U$. Assuming that $T$ and $U$ have been instrumented using distinct symbols $i$ and $j$, yielding $\instrument{T}{i}$ and $\instrument{U}{j}$, the sequential program operates as follows. First, the read, write and current locksets for $\instrument{T}{i}$ and $\instrument{U}{j}$ are initialized to be empty, and for each shared variable $s$, the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ are initialized to the full set of mutexes, $\mutexes$.  The main procedures of the instrumented thread templates, $\mainforthread{\instrument{T}{i}}$ and $\mainforthread{\instrument{U}{j}}$, are then executed in turn (the order does not matter, due to renaming). Finally, an assertion checks for consistent use of mutexes: if $s$ is written during execution of $\instrument{T}{i}$ and accessed during execution of $\instrument{U}{j}$, or vice-versa, then the locksets $\lockset{s}{i}$ and $\lockset{s}{j}$ must contain at least one common mutex.

\begin{figure}
\begin{tabular}{l}
$\currentlockset{i} = \emptyset;$ $\hasbeenread{i} = \emptyset;$ $\hasbeenwritten{i} = \emptyset;$ \\
$\currentlockset{j} = \emptyset;$ $\hasbeenread{j} = \emptyset;$ $\hasbeenwritten{j} = \emptyset;$ \\
\texttt{for} $s \in \sharedvars$ \texttt{do} $\lockset{s}{i} = \mutexes;$ $\lockset{s}{j} = \mutexes;$ \medskip
\\

@@

\medskip\noindent\textbf{Soundness }
%
We sketch an argument that if the program of Figure~\ref{fig:sequentialization} is correct, it is impossible for a thread executing template $T$ to race with a thread executing template $U$, under the assumption that the threads are guaranteed to terminate. Let us assume that the program of Figure~\ref{fig:sequentialization} is correct, and suppose (by way of contradiction) that a thread executing $T$ can in fact race with a thread executing $U$, on some shared variable $s$.  By our hypothesis that the program of Figure~\ref{fig:sequentialization} is correct, and that the threads terminate, the assertion checked at the end of the program guarantees at least one mutex, say $m$, belongs to both $\lockset{s}{i}$ and $\lockset{s}{j}$.  By the definition of a lockset (and according to the manner in which shared accesses are instrumented in Figure~\ref{fig:instrumentation}), this means that $m$ is held during every access to $s$ by both $\instrument{T}{i}$ and $\instrument{U}{j}$. As a result, $m$ must be unlocked and locked between the two accesses, which contradicts that the pair of accesses is racing.

We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}. In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.

@@

\medskip\noindent\textbf{Architecture }
%
Figure~\ref{fig:whoop} depicts the \whoop toolchain. Initially, \whoop accepts a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver and allow it to be subsequently analyzed for data races. Both the driver and the Linux model are passed to the translation phase of \whoop (see Figure~\ref{fig:whoop} -- A), which uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the original driver into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral).

Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

We engineered the Chauffeur and \whoop components of our toolchain (denoted with grey boxes in Figure~\ref{fig:whoop}).  For the remaining components, we were able to re-use industrial-strength tools that are robust and battle-proven via their use in many complex software projects.

\medskip\noindent\textbf{Extracting Entry Point Information }
%
Chauffeur is a Clang-based frontend that traverses the abstract syntax tree (AST) of the original driver (using a Clang visitor pass) and extracts all entry point identifier names, together with the identifier names of their corresponding kernel API functions. Linux drivers define entry points in a standard way (see Figure~\ref{fig:entrypoints} for an example of how the generic\_nvram driver defines the entry points for the \texttt{file\_operations} API); Chauffeur identifies these definitions in the AST and outputs the relevant information in an XML file, which is subsequently parsed by \whoop and used during the instrumentation.

\begin{figure}[t]
\begin{lstlisting}
const struct file_operations nvram_fops = {
  .llseek = nvram_llseek,
  .read = read_nvram,
  .write = write_nvram,
  .unlocked_ioctl = nvram_unlocked_ioctl,
};
\end{lstlisting}
\caption{Entry points definitions in the generic\_nvram driver}
\label{fig:entrypoints}
\end{figure}

\medskip\noindent\textbf{Translation for Verification }
%
Next, the driver is compiled by Clang to LLVM-IR~\cite{lattner2004llvm}, a low-level assembly-like language in single static assignment (SSA) form. Function calls (e.g.\ for locking and unlocking) are preserved in this translation, which is useful for our lockset instrumentation.

SMACK then translates the driver from LLVM-IR to Boogie IVL, which is the input language of \whoop. An important feature of SMACK is that it leverages the pointer-alias analyses of LLVM to efficiently model the heap manipulation operations of C programs in Boogie IVL. This means that \whoop does not need to directly deal with pointers and alias analysis, a hard problem on its own, allowing us to reuse robust existing techniques and focus instead on verification efforts.

To achieve scalability, SMACK uses a \emph{split} memory model that exploits an alias analysis to soundly partition memory locations into non-overlapping equivalence classes that do not alias. This has been shown to lead to more tractable verification compared with a \emph{monolithic} model where the heap is considered to be an array of bytes~\cite{rakamaric2009scalable}. The split memory model is based on \emph{memory regions}, which are maps of integers that model the heap. A benefit of using this model is that distinct memory regions denote disjoint sections of the heap. We leverage this knowledge inside \whoop to guide and optimize our lockset instrumentation and analysis, and to create a fine-grained context-switch instrumentation as discussed in \S\ref{corral}. Note that the SMACK's memory model can potentially be unsound in (typically rare) situations where programs use unaligned byte-level memory accesses. The latest release of SMACK addresses this issue, and as future work we plan to integrate the latest SMACK with \whoop to further improve soundness.

\medskip\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (on the Boogie IVL source code of each entry point) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation, is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentation and optimization passes.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. However, it is relatively easy to enhance our tool with knowledge of other locking primitives. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. The are two reasons behind this: (i) it is arguably hard to detect such locks using static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite{corbet2005linux}.

\medskip\noindent\textbf{Watchdog Race-Checking Instrumentation }
%
Data race detection is performed by introducing sets containing the locks that are consistently held by each shared variable and sets containing all shared variables that are read and written (see \S\ref{sec:symbolicpairwise} and the instrumentation of Figure~\ref{fig:instrumentation}). These sets can be modeled directly in Boogie as characteristic functions, using maps. However, this requires the use of quantifiers to express properties related to set contents.  For instance, to express that a set $X$ of elements of type $A$ is empty, where $X$ is represented as a map from $A$ to $\mathsf{Bool}$, we would require the quantified expression $\forall a : A : \neg X[a]$.  It is well known that automated theorem proving in the presence of quantified constraints is challenging, and that theorem provers such as Z3~\cite{de2008z3} are often much more effective when quantifiers are avoided.

@@

\medskip\noindent\textbf{Generating Loop and Procedure Summaries }
%
Early in the development of \whoop we experimented with reasoning about recursion-free drivers by resorting to full inlining.  We found that this did not scale to large drivers: when we tried to apply \whoop to the r8169 ethernet driver (see \S\ref{evaluation}), we quickly exhausted the memory limits of our experimental platform as the driver exhibited recursion, which cannot be fully inlined and blows up.
